\documentclass{llncs}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc}


%\usepackage{url}
\usepackage{multirow}
\usepackage{listings}
\usepackage{amsmath}  % for equation*
\usepackage{array}    % for tabular
\usepackage{verbatim} % for comment
\usepackage{wrapfig}
\usepackage[final]{microtype}
\usepackage[pdfborder={0 0 0}]{hyperref}

\newcommand\forAll[1]{\forall \, #1 \, . \,}
\newcommand\forAllII[2]{\forall \, #1 \, #2 \, . \,}

\newcommand\propno[1]{(\emph{#1})}
\newcommand\hs[1]{\texttt{#1}}

% \raggedbottom

\lstnewenvironment{code}[1][]
  {\noindent
   \vspace{-0.5\baselineskip}
   \lstset{basicstyle=\ttfamily,
           frame=single,
           language=Haskell,
           keywordstyle=\color{black},
           #1}
   \fontsize{8pt}{8pt}\selectfont}
  {}

\newcommand\NOTE[1]{} % \mbox{}\marginpar{\fontsize{8pt}{8pt}\selectfont\raggedright\hspace{0pt}\emph{#1}}}

\def\maindocument{} % To tell tikz images that they are not stand alone

\begin{document}

\title{Integrating Theory Exploration in a Proof Assistant}

\author{Moa Johansson \and Dan Ros\'en \and Nicholas Smallbone}
\institute{Department of Computer Science and Engineering, Chalmers University of Technology
	\email{\{moa.johansson,nicsma,danr\}@chalmers.se}
	}

\authorrunning{Johansson, Ros\'en and Smallbone}
\titlerunning{Integrating Theory Exploration in a Proof Assistant}

\maketitle

\begin{abstract}
Theory exploration is a technique for automatically discovering novel and interesting lemmas in a given mathematical theory development. This paper describes Hipster, a system integrating theory exploration with the proof assistant Isabelle with the intention of increasing automation for inductive proofs. 
Existing Isabelle-tools like Sledgehammer can automatically pick from the existing lemma library those that are required for automation of many proofs. Hipster is an extension of this approach which also allows for discovery of new or missing lemmas which require induction, thus increasing the number of proofs that can be automated.
Hipster can be used in two main ways, either in exploratory mode to automatically generate basic lemmas about a given set of function symbols in a new theory development, or in proof mode for a particular proof attempt, trying to discover lemmas that would allow the current goal to be proved. 

\end{abstract}

\section{Introduction}
%\input{Introduction}

What is theory exploration an why we want it in an interactive prover. 

Which is because we can improve on Sledgehammer-style interactions in cases when the required lemmas simply are not available, so Sledgehammer would fail. 

We improve on theory exploration like HipSpec by allowing the user more control, of for instance which symbols that gets sent to the exploration system, i.e. if we have a very large theory the user chooses suitable subsets of functions to explore together.

A third contribution is the Blanchette variables, as Isabelle and Haskell deal with partial functions and their semantics in different ways.  

By integrating it in Isabelle, we add a new layer on top of Haskell-HipSpec which allows us to record and store results for later use.

\section{Background}
Short description of HipSpec to go in here.

\section{Hipster}
This section shows a bunch of examples and explains how it all works.

% Dealing with Partial Functions
\section{Dealing With Partial Functions}
\input{PartialFunctions}

\section{Evaluation}
Some form of evaluation????

\section{Related Work}
%\input{RelatedWork}

\section{Conclusion and Further Work}
%\input{Conclusion}

\bibliographystyle{plain}
\bibliography{bibfile}

\end{document}
