\section{Further Work}
\label{sec:further}
The discovery process is currently limited to equational lemmas. We
plan to extend the theory exploration to also take side conditions
into account. If theory exploration is called in the middle of a proof
attempt, there may be assumptions associated with the current
sub-goal, which could be a useful source of side conditions. For
example, if we are proving a lemma about sorting, there will most
likely be an assumption involving the ``less than'' operator; this
suggests that we should look for equations that have ``less than'' in
a side condition. Once we have a candidate set of side conditions, it
is easy to extend QuickSpec to find equations that assume those conditions.

The parameters for Hipster, e.g. the number of QuickSpec generated tests, the runtime for Metis and so on, are largely based on heuristics from development and previous experience. There is probably room for fine-tuning these heuristics and possibly adapt them to the theory we are working in. We plan to add and experiment with additional automated tactics in Hipster. Again, we expect that different tactics will suit different theories.
 
Another interesting area of further work in theory exploration is reasoning by analogy. In the example in section \ref{sec:tree}, theory exploration discovers lemmas about \texttt{mirror} and \texttt{tmap} which are analogous to lemmas about lists and the functions \texttt{rev} and \texttt{map}. Machine learning techniques can be used to identify similar lemmas \cite{acl2ml}, and this information could then be used to for instance suggest new combinations of functions to explore, new connections between theories or directly suggest additional lemmas about trees by analogy to those on lists.

\section{Summary}
\label{sec:concl}

Hipster integrates lemma discovery by theory exploration in the proof
assistant Isabelle/HOL. We demonstrated two typical use cases of how this
can help and speed up theory development: by generating interesting
basic lemmas in \emph{exploration mode} or as a lemma suggestion
mechanism for a stuck proof attempt in \emph{proof mode}. The user can
control what is discovered by varying the background theory, and by
varying Hipster's ``routine reasoning'' and ``difficult reasoning''
tactics; only lemmas that need difficult reasoning (e.g. induction)
are presented.

Hipster complements tools like Sledgehammer: by discovering missing
lemmas, more proofs can be tackled automatically. Hipster succeeds in
automating inductive proofs and lemma discovery for small, but
non-trivial, equational Isabelle/HOL theories. The next step is to
increase its scope, to conditional equations and to bigger theories:
our goal is practical automated inductive proof for Isabelle/HOL.

%Conditionals, analogy between different datastructures when similar lemmas has been found about them. 
